Inferring community characteristics in labelled networks

### 2021-05-16 nips/inference.tex
Split into multiple tex files

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2021

% ready for submission
\PassOptionsToPackage{numbers, compress}{natbib}
\usepackage[final]{neurips_2021}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2021}

% to compile a camera-ready version, add the [final] option, e.g.:
%\usepackage[final]{neurips_2021}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2021}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bbm}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}

%images
\graphicspath{{./img/}}

%bibliography
\bibliographystyle{plainnat}

\title{Inferring community characteristics in labelled networks}
% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{
  Lawrence Tray \\
  Department of Engineering \\
  University of Cambridge \\
  \texttt{lpt30@cam.ac.uk} \\
  % examples of more authors
  \And
  Ioannis Kontoyiannis \\
  Department of Mathematics \\
  University of Cambridge \\
  \texttt{ik355@cam.ac.uk}
}

%custom commands
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Bcal}{\mathcal{B}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\specialchoose}{\genfrac{\{}{\}}{0pt}{}}
\newcommand{\Expect}{\mathbb{E}}

%envs
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\begin{document}

\maketitle

\begin{abstract}
We present a novel generative model for describing vertex-labelled undirected graphs. This is a two-layered model. First a probabilistic mapping from vertex labels to latent block memberships and from these block memberships we generate a graph according to the micro-canonical stochastic block model.

With this framework we are able to efficiently sample the parameters of the vertex label to block membership mapping. This allows us to automatically determine which features have the largest impact on graphical structure.
\end{abstract}

\section{Introduction}

There is a wealth of networks in the world. Some examples of this graphical data are social networks, website hyperlinks and academic collaboration with more being produced each second. It is clear we need tools to analyse this increasingly omni-present form of data.

A somewhat surprising property of real-world graphs is that they exhibit strong community structure. In other words, each node will often belong to a cluster of densely connected nodes. This property is often exploited by graph compression algorithms and there is high interest in recovering the communities from the observed graph.

A common subset of graphical data is the labelled network. This is a graph where we have information about the properties of each node. We shall refer to these node properties as features. One of the most common questions we can ask of this dataset is what features have the largest impact on the structure of the graph. For example, when analysing an academic collaboration graph, one may wish to ask what impact gender has on the structure of the network.  Nevertheless, this analysis is often vulnerable to confounding variables. While gender very well may impact the structure of the graph, there is often a better explanatory variable for the structure.

There is space to bridge the gap between these two approaches. Rather than determining whether a feature impacts the graphical structure directly, we can use the community concept as a stepping stone in our analysis. Therefore, we extract which features have the largest impact on overall graphical structure. This can be thought of as a form of dimensionality-reduction where we only keep the node features that are useful in predicting community memberships.

\section{Preliminaries}

This section defines some preliminary concepts required for the subsequent analysis. We first need a model for community-like structure in a graphical network. For this we adopt the stochastic block model (SBM) - widely used across academia. The premise is that each node in the graph belongs to a unique community called a block. The probability that two nodes are connected depends only on the block memberships of each. Graphs drawn from this ensemble exhibit community structure. Specifically, we will use the microcanonical variant of the SBM, proposed by \citet{Peixoto-Bayesian-Microcanonical}. A paraphrased definition is given below.

\begin{definition}[Microcanonical NDC-SBM]
	\label{defn:microcan-ndc-sbm}
	Let $N \in \Integers^{+}$ denote the number of vertices in an undirected graph. The block memberships are encoded by a vector $b$ of length $N$ where each entry $b_i \in \{1, 2 \dots B\}$. $B \in \Integers^{+}$ is the number of non-empty blocks. Let $e$ be a $B \times B$ matrix of edge counts between blocks ($e_{rs}$ is number of edges from block $r$ onto block $s$ - or twice that number if $r=s$). For undirected graphs $e$ is symmetric. For a non-degree-corrected stochastic block model (NDC-SBM), we say that the graph $A$ is generated as follows:
	%
	\begin{equation}
		A \sim \textrm{NDC-SBM}_{\textrm{MC}} (b, e)
	\end{equation}
	%
	Where edges are placed uniformly at random but respecting the constraint imposed by $e$ and $b$. The additional parameters $N$ and $B$ are omitted as they are inferred from the shapes of $b$ and $e$. If we interpret $A$ as an adjacency matrix, then this constraint can be written formally as: $e_{rs} = \sum_{i,j} A_{ij} \one \{b_i = r\} \one \{b_j = s\}$.
\end{definition}

Nevertheless, this formulation does not accept high degree variability within blocks as is typical of real-world data. Indeed, the NDC-SBM favours a partition into high-degree and low-degree nodes rather than clusters of inter-connected nodes. We therefore introduce the degree-corrected SBM (DC-SBM) to circumvent these issues.

\begin{definition}[Microcanonical DC-SBM]
	\label{defn:microcan-dc-sbm}
	This is much like the NDC-SBM but has an additional parameter $k$ which is an $N$-length vector encoding the degree sequence ($k_i$ is the degree of vertex $i$). Therefore, we write:
	%
	\begin{equation}
		A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, e, k)
	\end{equation}
	%
	Once again, edges are placed uniformly at random but respecting the constraints imposed by the parameters. The DC-SBM has the additional constraint that $k_i = \sum_{j} A_{ij}$. In what follows, we will always assume the degree-corrected model unless otherwise specified.

\end{definition}

\section{Latent block generative model}

In this section we propose a novel generative model for modelling labelled networks. We restrict our analysis to labelled, undirected graphs with $N$ nodes. We define the vector $x_i \in \Xcal^D$ as the feature vector for the $i$'th vertex. Each vertex has $D$ total features and we assume all entries take values from the same set $\Xcal$. For the majority of datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. The feature vectors $\{x_i\}_{i=1}^{N}$ are subsumed into the $N \times D$ matrix $X$.

The proposed generative model (which we call the latent block generative model - LBGM) is given in figure \ref{fig:lbgm}. We start, with the feature matrix X and generate a vector of block memberships $b$. The parameters of this step are encapsulated by $\theta$. Each feature vector $x_i$ is treated independently and used to generate the block membership $b_i$. Each entry $b_i \in \{1, 2 \dots B\}$ where $B$ is the number of blocks and a constant in our model. We choose a single softmax layer to model $p(b_i | x_i, \theta)$. More complex models are possible but then deriving meaning from inferred parameter distributions is more difficult. Summarising, we write $p(b | X, \theta)$ as follows:
%
\begin{equation}
	p(b| X, \theta) = \prod_{i=1}^{N} p(b_i | x_i, \theta) = \prod_{i=1}^{N} \phi_{b_i} (x_i; \theta)
	= \prod_{i=1}^{N} \frac{\exp\left(w_{b_i}^T \tilde{x}_i\right)}{\sum_{k=1}^{B} \exp \left( w_k^T \tilde{x}_i\right)}
\end{equation}
%
Where $\tilde{x} \coloneqq \left[x_1, x_2, \dots x_D, 1\right]^T$ is an augmented version of $x$ that allows for a bias term. The parameters for this stage are denoted $\theta$ and consist of the all $B$ weight vectors $\theta = \{w_k\}_{k=1}^{B}$. Each $w_k$ has dimension $D+1$. We could instead write the parameters $\theta$ as a $B \times (D+1)$ matrix of weights $W$; this form has use computationally as then $Z_i = W \tilde{x}_i$ is the input to the softmax activation function.

\begin{figure}[!h]
	\centering
	\begin{tikzpicture}[
		roundnode/.style={circle, draw=black, minimum size=7mm},
		squarednode/.style={rectangle, draw=black, minimum size=7mm}
		]
		% nodes
		\node[roundnode] (X) at (0, 0) {$X$};
		\node[squarednode] (b) at (3, 0) {$b$};
		\node[roundnode] (A) at (6, 0) {$A$};

		% arrows
		\draw[->] (X.east) -- node[above] {$\theta$} (b.west);
		\draw[->] (b.east) -- node[above] {$\psi$}(A.west);
	\end{tikzpicture}
	\caption{Latent block generative model}
	\label{fig:lbgm}
\end{figure}

With the block memberships $b$ generated, we then draw the graph $A$ from the microcanical DC-SBM (\ref{defn:microcan-dc-sbm}) with additional parameters $\psi = \{e, k\}$. In a slight abuse of notation we denote the inter-block edge count matrix with $e = \psi_e$ and the degree sequence $k = \psi_k$ to make explicit that these parameters are contained in $\psi$.
%
\begin{equation}
	A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi_e, \psi_k)
\end{equation}

\subsection{Prior selection}

Before performing any inference, we must specify priors on $\theta$ and $\psi$. For $\theta$ it seems sensible to choose a Gaussian prior, with zero mean and variance matrix $\sigma^2_\theta I$ such that each element of $\theta$ is independent and distributed like $\sim \Gaussian(0, \sigma_\theta^2)$. In vector form, the prior for $\theta$ is therefore:
%
\begin{equation}
	p(\theta) = \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right)
	\label{eqn:theta-prior}
\end{equation}
%
We will see that this form of prior is equivalent to a regularisation term in neural network training that penalises extreme weight magnitudes. For $\sigma_\theta^2 \rightarrow \infty$ this becomes an uninformative uniform prior.

In our model, the block memberships vector $b$ is an intermediate latent variable and so we are not free to choose a prior for it. Nevertheless, as far as inference on the right-hand-side of figure \ref{fig:lbgm}, we regard $p(b | X)$ as a pseudo-prior on $b$. We can show (appendix \ref{appdx:b|x}) that our choice of prior for $p(\theta)$ in equation \ref{eqn:theta-prior} leads to a uniform $p(b | X)$ in equation \ref{eqn:b-pseudo-prior}.
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is a uniform distribution that only depends on the number of blocks $B$. This simplifies the problem nicely as evaluating $p(b | X)$ does not require knowing the exact value $X$ takes. \citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for the additional microcanonical SBM parameters $\psi$ which we adopt. Peixoto's idea is to write the joint prior on $(b, e, k)$ as a product of conditionals $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. For our purposes we must insert a conditioning on $X$, to form our pseudo-prior for $b$ and $\psi$, to give equation \ref{eqn:joint-pseudo-prior}.
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b)
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
Where the simplification is made apparent by noting $(\psi \indep X) | b$. We then borrow the priors proposed by \citet{Peixoto-Bayesian-Microcanonical} for $p(\psi | b)$, repeated here for reference.
%
\begin{equation}
	p(\psi | b) = p(e | b) p(k | e, b) = \left[ \specialchoose{ \specialchoose{B}{2} }{ E} \right]^{-1} 
	\cdot \left[ \prod_r \frac{\prod_j \eta_j^r !}{n_r! q(e_r, n_r)} \right]
\end{equation}
%
Where $\specialchoose{n}{m}$ is shorthand for $\binom{n+m-1}{m} = \frac{(n+m-1)!}{(n-1)!(m)!}$ which can be thought of as the total number of distinct histograms with $n$ bins under the constraint they sum to $m$. $E = \frac{1}{2} \sum_{r,s} e_{rs}$ is the total number of edges in the graph. Importantly, $E$ is not allowed to vary and so $p(e|b)$ is uniform with respect to $e$. The variable $\eta_j^r$ is introduced to denote the number of vertices in block $r$ that have degree $j$. Formally, $\eta_j^r \coloneqq \sum_{i} \one\left\{b_i = r \right\} \one \left\{k_i = j \right\}$. Furthermore, $q(m, n)$ is the number of different histograms with at most $n$ non-zero bins that sum to $m$. $q(m, n)$ is related to but different from $\specialchoose{n}{m}$. Lastly, $e_r \coloneqq \sum_{s} e_{rs}$ is the total number of half edges in block r and $n_r \coloneqq \sum_{i} \one\{b_i = r\}$ is the number of vertices assigned to block $r$. Importantly, we have computable forms for $p(\theta)$ and $p(b, \psi | X)$ which will be useful for performing inference.

@@

\citet{Peixoto-MCMC} proposes a Monte Carlo method which we will base our approach on. It relies on writing the posterior in the following form:
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b)
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised density we wish to sample from. In other words, we wish to construct a Markov chain that has $\pi_b(\cdot)$ as its invariant distribution. We can break $\pi_b$ down as follows:
%
\begin{align*}
	\pi_b(b) &= p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	&= p(b|X) p(A, \psi^* | b, X) \\
	&= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X)
\end{align*}
%
Since we are using the microcanonical SBM formulation, there is only one value of $\psi$ that is compatible with the given $(A, b)$ pair. We denote this value $\psi^* = \{k^*, e^*\}$. Specifically, $k^*_i = \sum_j A_{ij} $ and $e^*_{rs} = \sum_{i, j} A_{ij} \one \{b_i=r\} \one\{b_j=s\}$. Therefore, the summation over all $\psi$ reduces to just the single $\psi^*$ term. We also define the microcanonical entropy of the configuration as.
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \Big( \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \Big)
\end{equation}
%
This entropy can be thought of as the description length of the graph because it is the sum of the information required to represent the graph given the parameters and the amount of information required to store the parameters (given the feature matrix $X$). The exact, from of the proposal distribution and accept-reject step is explored throughly by \citet{Peixoto-MCMC}. There is a widely available library, \verb*|graph-tool| \cite{peixoto_graph-tool_2014}, implementing this algorithm. The only modification we make is in the block membership prior $p(b)$ which we replace with $p(b|X)=B^{-N}$ which is a uniform distribution and so cancels out in the MH accept-reject step.

@@

\section{Experiments}

We apply the developed methods to a variety of datasets, to show the power of the method across different scenarios. We choose a range of node counts $N$, edge counts $E$ and feature space dimension $D$.

\begin{itemize}
	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) - network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the course of the day. Vertex features include class membership, gender and whether or not the student is a teacher or pupil. These data were collected on consecutive days in October 2009. We choose to analyse just the second day. 
	\item \textbf{College football teams} \cite{Evans_2010_football} - network of American college football teams and their interactions. Vertex labels are the division a team belongs to.
	\item \textbf{Law firm} \cite{LawFirm} - a network of relationships between members of a law firm. Each relationship is categorised according to type: coworkers, friends or advice.
	\item \textbf{Maier Facebook Egonet} \cite{FB-Maier} - network of the author's Facebook friends list. Each vertex has been manually labelled with a variety of features describing their relationship to the author.
	\item \textbf{Twitch users} \cite{twitch} - a network of user-user friendships on the streaming service Twitch. Vertex labels are extracted according to video-games played, location and streaming habits. This dataset is also broken down into disjoint networks according to language. We only consider the English users with is a subnet with $N=7126$ vertices and $E=35324$ edges).
\end{itemize}

A table summarising the results is given in table \ref{tab:results}.

\begin{table}[]
	\centering
	\begin{tabular}{c|ccc|c|cc}
		Dataset & $N$ & $E$ & $D$ & $B$ & $\bar{S}_b /N$ & $\bar{U}_\theta /N$ \\ \hline
		Primary school & 238 & 5539 & 13 &  18 & 43.0  & 2.75 \\
		Maier FB egonet &     &     &     &   &           &             \\
		&     &     &     &              &     &       
	\end{tabular}
\end{table}

\subsection{Primary school dynamic contacts}

These data were originally collected by \citet{schools} to quantify the transmission opportunities for respiratory infections within a primary school context (ages 6-11 in France). However, we seek to ask a simpler question. What features best describe how people interact with one another in a primary school context. The only vertex features we have available are school-class (one of 10 values - 2 per year group), gender and a distinction between teachers and pupils.

We must first choose the number of blocks to consider $B$ to define the coarseness of our analysis. A total of 10 classes would suggest that $B=10$ is a natural starting point. With this value of $B$ we first sample from the block membership posterior $b^{(t)} \sim p(b | A, X)$. We visualise the inferred partition in figure \ref{fig:school-graph}

\begin{figure}[!h]
	\centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-graph.png}
		\caption{Softmax weighting samples for each block index}
		\label{fig:school-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-null.png}
		\caption{}
		\label{fig:school-null}
	\end{subfigure}
	\caption{Primary school dynamic contacts network \cite{schools}}
\end{figure}

\section{Extensions}

\section{Conclusion}

\begin{ack}
Use unnumbered first level headings for the acknowledgments. All acknowledgments
go at the end of the paper before the list of references. Moreover, you are required to declare
funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2021/PaperInformation/FundingDisclosure}.

Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to automatically hide this section in the anonymized submission.
\end{ack}

%references
\bibliography{sources.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Checklist}

%%% BEGIN INSTRUCTIONS %%%
The checklist follows the references.  Please
read the checklist guidelines carefully for information on how to answer these
questions.  For each question, change the default \answerTODO{} to \answerYes{},
\answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
justification to your answer}, either by referencing the appropriate section of
your paper or providing a brief inline description.  For example:
\begin{itemize}
  \item Did you include the license to the code and datasets? \answerYes{See Section gen-inst.}
  \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
  \item Did you include the license to the code and datasets? \answerNA{}
\end{itemize}
Please do not modify the questions and only use the provided macros for your
answers.  Note that the Checklist section does not count towards the page
limit.  In your paper, please delete this instructions block and only keep the
Checklist section heading above along with the questions/answers below.
%%% END INSTRUCTIONS %%%

\begin{enumerate}

\item For all authors...
\begin{enumerate}
  \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
    \answerTODO{}
  \item Did you describe the limitations of your work?
    \answerTODO{}
  \item Did you discuss any potential negative societal impacts of your work?
    \answerTODO{}
  \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
    \answerTODO{}
\end{enumerate}

\item If you are including theoretical results...
\begin{enumerate}
  \item Did you state the full set of assumptions of all theoretical results?
    \answerTODO{}
	\item Did you include complete proofs of all theoretical results?
    \answerTODO{}
\end{enumerate}

\item If you ran experiments...
\begin{enumerate}
  \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
    \answerTODO{}
  \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
    \answerTODO{}
	\item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
    \answerTODO{}
	\item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
    \answerTODO{}
\end{enumerate}

\item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
\begin{enumerate}
  \item If your work uses existing assets, did you cite the creators?
    \answerTODO{}
  \item Did you mention the license of the assets?
    \answerTODO{}
  \item Did you include any new assets either in the supplemental material or as a URL?
    \answerTODO{}
  \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
    \answerTODO{}
  \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
    \answerTODO{}
\end{enumerate}

\item If you used crowdsourcing or conducted research with human subjects...
\begin{enumerate}
  \item Did you include the full text of instructions given to participants and screenshots, if applicable?
    \answerTODO{}
  \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
    \answerTODO{}
  \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
    \answerTODO{}
\end{enumerate}

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Appendix}

\subsection{Derivation of conditional block distribution given feature matrix}
\label{appdx:b|x}

We wish to determine the form of $p(b| X)$. This can be done by integrating over the joint probability with respect to $\theta$.
%
\begin{align*}
	p(b | X) &= \int p(b , \theta| X, \theta) d\theta = \int p(b | X, \theta) p(\theta | X) d\theta \\
	&=\int p(b | X, \theta) p(\theta) d\theta = \int \prod_{i=1}^{N} \phi_{b_i}(x_i; \theta) p(\theta) d\theta \\
	&= \prod_{i=1}^{N} \int \frac{\exp(w_{b_i}^T \tilde{x}_i) \prod_{j=1}^{B} \Gaussian(w_j; 0, \sigma_\theta^2 I)}{\sum_{k=1}^{B} \exp(w_{k}^T \tilde{x}_i)} dw_{1:B}
\end{align*}
%
We note that $b_i \in {1, 2, \dots B}$ and so the integral's value is unchanged with respect to $b_i$. The integrand has the same form no matter which value $b_i$ takes as the prior is the same for each $w_j$. As such the integral can only be a function of at most $\tilde{x}_i$ and $\sigma_\theta^2$ as it is symmetric with respect to $b_i$ and all the various $w_j$ are integrated out as they are dummy variables. Therefore, denoting the integral by the (unknown) function $f(\tilde{x}_i, \sigma_\theta^2)$, we write $p(b| X)$ as follows:
%
\begin{align*}
	p(b | X) &= \prod_{i=1}^{N} f(\tilde{x}_i, \sigma_\theta^2) = \textrm{const w.r.t } b = c
\end{align*}
%
As this is a constant with respect to $b$ we conclude that $p(b | X)$ must be a uniform distribution. $\nicefrac{1}{c}$ is simply the size of the set of values that $b$ can take. We know $b_i \in \mathcal{B} = \{1, 2, \dots B\}$. Therefore, $b \in \mathcal{B}^N$ and $|\mathcal{B}^N| = |\mathcal{B}|^N = B^N= \nicefrac{1}{c}$. Putting this all together we show that:
%
\begin{equation}
	p(b | X) = B^{-N}
\end{equation}

\subsection{Derivation of U gradient with respect to feature parameters}
\label{appdx:gradu}
The goal is to determine $\nabla U(\theta)$, the gradient of the negative log posterior with respect to the parameters. We repeat the form of $U(\theta)$ in equation \ref{eqn:U-form-appdx}.
%
\begin{equation}
	U(\theta) = \left( \sum_{i=1}^{N} \sum_{j=1}^{B} y_{ij} \log \frac{1}{a_{ij}} \right)
	+ \frac{1}{2\sigma_\theta^2} ||\theta||^2
	\label{eqn:U-form-appdx}
\end{equation}
%
Where $y_{ij}$ is independent of $\theta$ and $a_{ij}$ is the output from the softmax layer, with form as given in equation \ref{eqn:a-ij}.
%
\begin{equation}
	a_{ij} \coloneqq \phi_{j} (x_i; \theta) = \frac{\exp(w_j^T \tilde{x}_i)}{\sum_{b=1}^{B} \exp(w_b^T \tilde{x}_i)}
	\label{eqn:a-ij} 
\end{equation}
%
We note that $\theta = \{w_k\}_{k=1}^B$, and as such we can write this in vector form $\theta = \left[w_1^T, w_2^T \dots w_B^T  \right]^T$. Therefore, $\nabla U(\theta) = \left[\nicefrac{\partial U}{\partial w_1}^T,\nicefrac{\partial U}{\partial w_2}^T \dots \nicefrac{\partial U}{\partial w_B}^T  \right]^T$; to compute $\nabla U(\theta)$ it suffices to find the form of $\nicefrac{\partial U}{\partial w_k}$ with respect to a general $k$.

To this end, we must first find partial derivatives of $a_{ij}$ and $||\theta||$ with respect to $w_k$. Starting with $a_{ij}$:
%
\begin{align}
	\frac{\partial a_{ij}}{\partial w_k} &= \frac
	{\tilde{x}_i \exp(w_j^T \tilde{x}_i) \delta_{jk} \cdot \sum_{b=1}^{B} \exp(w_b^T \tilde{x}_i) 
		- 
	\exp(w_j^T \tilde{x}_i) \cdot \tilde{x}_i \exp(w_k^T \tilde{x}_i)}
	{\left( \sum_{b=1}^{B} \exp(w_b^T \tilde{x}_i) \right)^2} \nonumber \\
	&= \tilde{x}_i \left( a_{ij} \delta_{jk} - a_{ij}a_{ik} \right) 
\end{align}
%
Where $\delta_{jk} \coloneqq \one \left\{ j = k \right\}$. Now moving onto the derivative of $||\theta||^2$:
%
\begin{equation}
	\frac{ \partial}{\partial w_k} ||\theta||^2 = \frac{\partial}{\partial w_k} \left( \sum_{b=1}^B ||w_b||^2 \right) = 2w_k
\end{equation}
%
We are ready to put this all together, to find the partial derivative of $U(\theta)$ with respect to each $w_k$:
\begin{align}
	\frac{\partial U}{\partial w_k} &= 
	\sum_{i=1}^{N} \sum_{j=1}^{B} y_{ij} 
	\left( \frac{-\tilde{x}_i}{a_{ij}} \left( a_{ij} \delta_{jk} - a_{ij} a_{ik} \right) \right)
	+ \frac{w_k}{\sigma_\theta^2} \nonumber \\
	&=  - \left( \sum_{i=1}^{N} \tilde{x}_i \left( y_{ik} - a_{ik} \sum_{j=1}^{B} y_{ij} \right)
	- \frac{w_k}{\sigma_\theta^2} \right) \nonumber \\
	&= - \left( \sum_{i=1}^{N} \Big\{ \tilde{x}_i (y_{ik} - a_{ik}) \Big\} - \frac{w_k}{\sigma_\theta^2} \right)
\end{align}
%
This is the required result. This form can be computed efficiently through matrix operations. The only property of $y_{ij}$ we have used in the derivation is the sum-to-one constraint $\sum_{j=1}^{B} y_{ij} = 1$ for all $i$.

\subsection{Choosing the MALA step-size}
\label{appdx:step-size}

For sampling from the $\theta$-chain of the block membership generator parameters, we employed the Metropolis Adjusted Langevin Algorithm (MALA). At iteration $t$, the proposed sample is generated by:
%
\begin{equation}
	\theta' = \theta^{(t)} - h_t \nabla U(\theta^{(t)}) + \sqrt{2h_t} \cdot \xi
\end{equation}
%
There are two competing objectives when choosing the step-size $h_t$. On the one hand, we want the step-size to be large so that we arrive at a high density region quickly. However, too large a step-size will lead to a lower acceptance ratio and thus inefficient sampling. A solution to this problem would be to slowly decrease the step-size with $t$ - often called simulated annealing. Therefore, we still have a short burn-in time but will not bounce around the mode for large $t$. As well as the trivial constraint for $h_t$ to be strictly positive and we introduce two further constraints as outlined by \citet{Bayesian-SGLD}:
%
\begin{equation}
	\sum_{t=1}^{\infty} h_t = \infty \qquad \textrm{and} \qquad
	\sum_{t=1}^{\infty} h_t^2 < \infty
	\label{eqn:h-constraints}
\end{equation}
%
The first constraint ensures that we have cover sufficient distance to arrive at any arbitrary point in our domain, no matter the starting point. The second constraint ensures that once we converge to the mode rather than simply bouncing around it. \citet{Bayesian-SGLD} propose the following form for a polynomially decaying step-size which we adopt:
%
\begin{equation}
	h_t = a(b + t)^{-\gamma}
\end{equation}
%
Where $a, b, \gamma$ are hyper-parameters to be chosen. We require $a,b > 0$ and $\gamma \in (0.5, 1]$ to satisfy equation \ref{eqn:h-constraints}. We find empirically that $a=250, b=1000, \gamma=0.8$ yield good burn-in times for our MALA implementations.

\end{document}

>>>>

@@

\citet{Peixoto-MCMC} proposes a Monte Carlo method which we will base our approach on. It relies on writing the posterior in the following form:
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b)
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised density we wish to sample from. In other words, we wish to construct a Markov chain that has $\pi_b(\cdot)$ as its invariant distribution. We can break $\pi_b$ down as follows:
%
\begin{align*}
	\pi_b(b) &= p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	&= p(b|X) p(A, \psi^* | b, X) \\
	&= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X)
\end{align*}
%
Since we are using the microcanonical SBM formulation, there is only one value of $\psi$ that is compatible with the given $(A, b)$ pair. We denote this value $\psi^* = \{k^*, e^*\}$. Specifically, $k^*_i = \sum_j A_{ij} $ and $e^*_{rs} = \sum_{i, j} A_{ij} \one \{b_i=r\} \one\{b_j=s\}$. Therefore, the summation over all $\psi$ reduces to just the single $\psi^*$ term. We also define the microcanonical entropy of the configuration as.
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \Big( \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \Big)
\end{equation}
%
This entropy can be thought of as the description length of the graph because it is the sum of the information required to represent the graph given the parameters and the amount of information required to store the parameters (given the feature matrix $X$). The exact, from of the proposal distribution and accept-reject step is explored throughly by \citet{Peixoto-MCMC}. There is a widely used library for Python made available under LGPL called \verb*|graph-tool| \cite{peixoto_graph-tool_2014}, which implements this algorithm. The only modification we make is in the block membership prior $p(b)$ which we replace with $p(b|X)=B^{-N}$ which is a uniform distribution and so cancels out in the MH accept-reject step.

@@
